---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sbar - Stage-based assessments in R 

<!-- badges: start -->
<!-- badges: end -->

sbar is a package that provides functions to allow the user to run various stage-based fisheries assessments in R. The stage-based assesments available are versions of the current (2020) Catch-Survey Analysis (CSA) found in the the NOAA Fish and Fisheries Toolbox and an implementation of a size-based delay-diffrenec model described in the throerteical paper by Schnute (1987). Model theory, implementation and testing are detailed in ........

## Installation
This package involves C++ code and the modelling framework of the package \code{TMB}.  Please ensure that you are able to compile source code and that this package is running smoothly on your device.

And sbar package can be installed from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("lbatts/sbar")
```
## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(sbar)
## basic example code
```
